Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-05 doc/thesis/objective.tex
Formatting for GPS

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by
\gls{ABC}~\autocite{rubin1984bayesianly,tavare1997inferring,fu1997estimating,beaumont2002approximate}.
All of the ingredients required to apply \gls{ABC} to this problem are readily
available. Simulating networks is straightforward under a variety of models.
Epidemics on those networks, and the corresponding transmission trees, can also
be easily simulated. As mentioned above, contact networks can profoundly affect
transmission tree shape, and those shapes can be compared using a highly
informative similarity measure called the ``tree
kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
\gls{SMC}, which has several advantages over other
algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
even more attractive approach~\autocite{del2012adaptive}. In summary, our
method to infer contact network parameters will combine the following:
stochastic simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
the term ``kernel-\gls{ABC}'' to refer to our method specifically.

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
schneeberger2004scale, clemencon2015statistical}, meaning that their degree
distributions follow a power law (although there has been some disagreement,
see \autocite{handcock2004likelihood, bansal2007individual}). Preferential
attachment has been postulated as a mechanism by which scale-free networks
could be generated~\autocite{barabasi1999emergence}. The \gls{BA}
model~\autocite{barabasi1999emergence} is one of the simplest preferential
attachment models, which makes it a natural choice to explore with our method.
The second aim of this work is to use simulations to investigate the parameters
of the \gls{BA} model, including whether they have a detectable impact on tree
shape, and whether they can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. At the time of this writing, there were $635400$ HIV
sequences publicly available in GenBank, annotated with 172 distinct countries
of origin. Since \gls{HIV} is almost always spread through either sexual
contact or sharing of injection drug supplies, the contact networks underlying
\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
be highly nonrandom~\autocite{clemencon2015statistical}. Moreover, since no
cure yet exists, efforts to curtail the progression of an epidemic have relied
on preventing further transmissions through measures such as \gls{tasp} and
education leading to behaviour change. The effectiveness of this type of
intervention can vary significantly based on the underlying structure of the
network and the particular nodes to whom the intervention is
targeted~\autocite{little2014using,wang2015targeting}. Due to this combination
of data availability and potential public health impact, \gls{HIV} is an
obvious context in which our method could be applied. Therefore, the third and
final aim of this work is to apply kernel-\gls{ABC} to fit the \gls{BA} model
to existing \gls{HIV} outbreaks.

>>>>

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by
\gls{ABC}~\autocite{rubin1984bayesianly,tavare1997inferring,fu1997estimating,beaumont2002approximate}.
All of the ingredients required to apply \gls{ABC} to this problem are readily
available. Simulating networks is straightforward under a variety of models.
Epidemics on those networks, and the corresponding transmission trees, can also
be easily simulated. As mentioned above, contact networks can profoundly affect
transmission tree shape, and those shapes can be compared using a highly
informative similarity measure called the ``tree
kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
\gls{SMC}, which has several advantages over other
algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
even more attractive approach~\autocite{del2012adaptive}. In summary, our
method to infer contact network parameters will combine the following:
stochastic simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
method is a type of kernel-\gls{ABC}~\autocite{nakagome2013kernel}. For ease of
exposition, we will often use the term ``kernel-\gls{ABC}'' to refer to our
method specifically.

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
schneeberger2004scale,clemenccon2015statistical}, meaning that their degree
distributions follow a power law (although there has been some disagreement,
see \autocite{handcock2004likelihood, bansal2007individual}). Preferential
attachment has been postulated as a mechanism by which scale-free networks
could be generated~\autocite{barabasi1999emergence}. The \gls{BA}
model~\autocite{barabasi1999emergence} is one of the simplest preferential
attachment models, which makes it a natural choice to explore with our method.
The second aim of this work is to use simulations to investigate the parameters
of the \gls{BA} model, including whether they have a detectable impact on tree
shape, and whether they can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. At the time of this writing, there were $635400$ HIV
sequences publicly available in GenBank, annotated with 172 distinct countries
of origin. Since \gls{HIV} is almost always spread through either sexual
contact or sharing of injection drug supplies, the contact networks underlying
\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
be highly nonrandom~\autocite{clemenccon2015statistical}. Moreover, since no
cure yet exists, efforts to curtail the progression of an epidemic have relied
on preventing further transmissions through measures such as \gls{tasp} and
education leading to behaviour change. The effectiveness of this type of
intervention can vary significantly based on the underlying structure of the
network and the particular nodes to whom the intervention is
targeted~\autocite{little2014using,wang2015targeting}. Due to this combination
of data availability and potential public health impact, \gls{HIV} is an
obvious context in which our method could be applied. Therefore, the third and
final aim of this work is to apply kernel-\gls{ABC} to fit the \gls{BA} model
to existing \gls{HIV} outbreaks.

### 2016-05-05 doc/thesis/preface.tex
Formatting for GPS

A version of chapter 2 has been submitted for publication in Virus Evolution
with the title ``Reconstructing network parameters from viral phylogenies.'' An
oral presentation entitled ``Phylodynamic inference of contact network
parameters with kernel-ABC'' was given based on chapter 2 to the 23rd HIV
Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA
(the presentation was delivered remotely). A poster based on chapter 2 entitled
``Likelihood-free estimation of contact network parameters from viral
phylogenies'' was presented at the Intelligent Systems for Molecular Biology
meeting in Orlando, Florida, USA.

>>>>

A version of chapter 2 has been submitted for publication with the title
``Reconstructing network parameters from viral phylogenies.'' An oral
presentation entitled ``Phylodynamic inference of contact network parameters
with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
Evolution meeting on April 25, in Woods Hole, Massachussets, USA (the
presentation was delivered remotely). A poster based on chapter 2 entitled
``Likelihood-free estimation of contact network parameters from viral
phylogenies'' was presented at the Intelligent Systems for Molecular Biology
meeting in Orlando, Florida, USA.
